The role of aneuploidy in the evolution of cancer drug resistance

### 2023-07-16 EvolutionaryRescue1.tex
yoav revised; updaed figs

Even evolutionary rescue occurs, it may take a long time; therefore, it is crucial to estimate the mean waiting time for rescue and the effect aneuploidy may have on it. 
We therefore calculate the mean time for the appearance of the first mutant that rescues the cell population.
This can occur either through the evolutionary trajectory $wildtype \rightarrow mutant$ or through the trajectory $wildtype \rightarrow aneuploid \rightarrow mutant$.
We start with the former.

Assuming no aneuploidy ($u=0$), we define $T_1$ to be the time at which the first mutant cell appears that will avoid extinction and will therefore rescue the population. Note that if extinction occurs, that is $m_{\infty}=0$, then it is implied that $T_1=\infty$, and vice versa if $T_1<\infty$ then $m_{\infty}>0$.

The number of successful mutants generated until time $t$ can be approximated by an inhomogeneous Poisson process with rate $R\left(t\right) = u p_a w_t$,
where $w_t=N\e^{\Delta_w t}$ is the number of wildtype cells at time $t$.
Note that 
\begin{equation}\label{integralR}
\int_0^{t}{R(z)\d z} = 
u p_a N \frac{\exp[{\Delta_w t}]-1}{\Delta_w} \approx 
u p_a N t,
\end{equation}
by integrating the exponential and because $\frac{\exp[\Delta_w t]-1}{\Delta_w}=\frac{1+\Delta_w t+O(t^2)-1}{\Delta_w}=t+O(t^2)$.
The probability density function of $T_1$ is thus
$R\left(t\right)\exp\left(-\int_0^{t}{R(z)\d z}\right)$. % TODO ref
Therefore, the probability density function of $(T_1 \mid T_1 < \infty)$ is
$f_1(t) = \frac{R\left(t\right)\exp\left(-\int_0^{t}{R(z)\d z}\right)}{p_{rescue}}$. 
\\

We are interested in the mean conditional time, $\tau_1=\mathbb{E}\left[T_1 \mid T_1<\infty\right]$, which is given by
\begin{equation}\label{meantime1}
\begin{aligned}
\tau_1 =
\int_{0}^{\infty}{t f_1(t) \d t} = 
\frac{\int_{0}^{\infty}{tR(t)\exp\left(-\int_0^{t}{R(z)\d z}\right) \d t}}{p_{rescue}} = 
\frac{\int_{0}^{\infty}{\exp\left(-\int_0^{t}{R(z)\d z}\right) \d t}}{p_{rescue}}
\end{aligned}
\end{equation}
after applying integration by parts.
Therefore, plugging \cref{integralR,rescue_prob2} in \cref{meantime1}, the probability density function of $T_1$ is given by:
\begin{align}
f_{T_1}\left(t_1\right)=R\left(t_1\right)\e^{-\int_0^{t_1} R(t)\d t}.
\end{align}
As a result, the time $T_1$ conditional on evolutionary rescue is given by:
\begin{align}
f_{T_1}\left(t_1|m_{t\rightarrow\infty}\neq0\right)=\frac{R\left(t_1\right)\e^{-\int_0^{t_1} R(t)\d t}}{1-\left(1-p_w\right)^N}.
\end{align}
The expectation of $T_1$ is:
\begin{align}\label{meantime1}
\tau_1=\mathbb{E}\left[T_1\right]&=\frac{\int_0^\infty\e^{-\int_0^\tau R(t)\d t}\d\tau}{1-\left(1-p_w\right)^N}=\frac{\int_0^\infty\e^{-uNp_a\frac{\e^{\Delta_w\tau}-1}{\Delta_w}}\,\d\tau}{1-\left(1-p_w\right)^N}.
\end{align}
The fration in the exponential of the integrand in \eqref{meantime1} can approximated as:
\begin{equation*}
\frac{\e^{\Delta_w\tau}-1}{\Delta_w}=\frac{1+\Delta_w\tau+O(\tau^2)-1}{\Delta_w}=\tau+O(\tau^2).
\end{equation*}
As a result, the mean time $\tau_1$ can be simplified:
\begin{align}\label{limitapprox}
\tau_1 =
\frac{\int_{0}^{\infty}{\exp\left(-u p_a N t\right) \d t}}{1-\e^{-Np_w}} \approx
\left(1+\e^{-Np_w}\right)\int_0^\infty\e^{-u p_a N t} \d t =
\frac{1+\e^{-Np_w}}{u p_a N} ,
\end{align}
where we use the approximation $\left(1-\e^{-Np_w}\right)^{-1}\approx1+\e^{-Np_w}$ and integrate the exponent.
\Cref{MeanTimeGrowthAneuploidyPlot} show the agreement between this approximating and simulation results for intermediate and large tumor sizes.

\textbf{THE REST OF THIS SECTION REQUIRES EDITING BY YOAV}

%Let $T=T_1+T_2$ be the time the first mutant cell that appears and rescues the population, where $T_1$ is the time at which the first aneuploid cell appears whose lineage will rescue the population and $T_2$ the time for this aneuploidy lineage to produce the first mutant that rescues the population. The expected mean time $\tau_1=\mathbb{E}\left[T_1\right]$ is identical to \eqref{meantime1} with adequate parameters.
%
%For an aneuploid cell, let $a_\tau$ be the number of its aneuploid descendents at time $\tau$. Each of these descentends have a probability $v\d \tau$ of producing a mutant which has a probability $p_m$ to survive. Consequently, the probability that the lineage descendent from a single aneuploid cell produces succesful mutant in the time interval $\left[\tau,\tau+\d \tau\right]$ is $vp_ma_t\d \tau$. Since the number of succesful mutants is a Poisson process, the probability of evolutionary rescue occurring until time $t$, given $a_\tau$, is given by:
%\begin{align}
%1-\exp\left(-\int_0^t vp_ma_\tau\d \tau\right)=1-\exp\left(-vp_mA_t\right),
%\end{align}
%where $A_t$ is the weight of the lineage until time $t$, defined as:
%\begin{equation}
%A_t=\int_0^t a_\tau\d \tau.
%\end{equation}
%As a result, to obtain $p_a\left(t\right)$ we average over all possible values of $A_t$:
%\begin{align}
%p_a\left(t\right)&=\int_0^\infty \d a P\left(A_T=a\right)\left[1-\e^{-vp_mA_t}\right]\\
%&=1-\mathbb{E}\left[\e^{-vp_mA_t}\right]
%\end{align}
%Following the steps outlines in \citep{weissman2009rate} we obtain:
%\begin{align}
%p_a\left(t\right)=\frac{\left(a_+-1\right)\left(1-a_-\right)\left(1-\e^{-\left(1-\delta\right)\left(a_+-a_-\right)t}\right)}{a_+-1+\left(1-a_-\right)\e^{-\left(1-\delta\right)\left(a_+-a_-\right)t}},
%\end{align}
%where
%\begin{align}
%a_\pm=\frac{2-\delta-vp_m\pm\sqrt{\left(\delta+vp_m\right)^2-4vp_m}}{2\left(1-\delta\right)}.
%\end{align}
%and
%\begin{align}
%\delta=\frac{\Delta_a}{\Delta_w}
%\end{align}
%
%The cumulative distribution of $T_2$ is given by $\frac{p_a\left(t\right)}{p_m}$ from which we obtain the mean time:
%\begin{align}
%\tau_2=\mathbb{E}\left[T_2\right]=\int_0^\infty\left(1-\frac{p_a\left(t\right)}{p_a}\right)\d t.
%\end{align}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
When $Nu\gg1$ the aneuploid population can be assumed to be deterministic and approximated by the solution to the system of ODEs:
\begin{equation}
a_t=\frac{Nu\e^{\Delta_wt}}{\Delta_w-\Delta_a}\left[1-\e^{\left(\Delta_w-\Delta_a\right)t}\right].
\end{equation}
As a result, when $N\gg1$ the number of successful mutants created by direct mutation or though aneuploidy are an inhomogeneous Poisson processes with the rates:

%\begin{align}
%f_{T_2}\left(t_2|m_{\infty}\neq0\right)=\frac{R_2\left(t_2\right)\e^{-\int_0^{t_2} R_2(t)\d t}}{1-\left(1-p_w\right)^N}
%\end{align}

%\begin{align}\label{meantime2}
%\tau_2=\mathbb{E}\left[T_2\right]&=\frac{\int_0^\infty\e^{-\int_0^\tau R_2(t)\d t}\d\tau}{1-\left(1-p_w\right)^N}=\frac{\int_0^\infty\e^{-uNp_a\frac{\e^{\Delta_w\tau}-1}{\Delta_w}}\,\d\tau}{1-\left(1-p_w\right)^N}.
%\end{align}
%
%\begin{align}\label{meantime3}
%\tau_3=\frac{\int_0^\infty\e^{-vNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}}\,\d\tau}{1-\left(1-p_w\right)^N}.
%\end{align}

\begin{align}
r_1\left(t\right)&=vp_m\int_0^ta_{\tau}\,\d\tau=\frac{uvNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_wt}-1}{\Delta_w}-\frac{\e^{\Delta_at}-1}{\Delta_a}\right),\\
r_2\left(t\right)&=vp_m\int_0^tw_{\tau}\,\d\tau=vNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}.
\end{align}
For large initial population sizes we can assume that both rescue mutations produced through direct mutation and aneuploidy are independent and, as a result, they can be merged into a single Poisson process with rate $\left(r_1+r_2\right)\left(t\right)$. Consequently, the mean time to the appearance of the first rescue mutant is: 
\begin{align}\label{meantimet2}
\tau_2=\frac{\int_0^\infty\e^{-\left(r_1+r_2\right)}\,\d\tau}{1-\left(1-p_w\right)^N}=\frac{\int_0^\infty\exp\left[-\frac{uvNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_w\tau}-1}{\Delta_w}-\frac{\e^{\Delta_a\tau}-1}{\Delta_a}\right)-vNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}\right]\,\d\tau}{1-\left(1-p_w\right)^N},
\end{align}
which we plot in \Cref{MeanTimeGrowthAneuploidyPlot} as a function of the initial population size.

We wish to obtain a simpler formula for $\tau_2$ in an analogous way to \eqref{limitapprox}. For this, we make use of the following expansions:
\begin{align*}
\frac{\e^{\Delta_w\tau}-1}{\Delta_w}&=\frac{1+\Delta_w\tau+\frac{\Delta_w^2\tau^2}{2}+O(\tau^3)-1}{\Delta_w}=\tau+\frac{\Delta_w}{2}\tau^2+O(\tau^3),\\
\frac{\e^{\Delta_a\tau}-1}{\Delta_a}&=\frac{1+\Delta_a\tau+\frac{\Delta_a^2\tau^2}{2}+O(\tau^3)-1}{\Delta_a}=\tau+\frac{\Delta_w}{2}\tau^2+O(\tau^3),
\end{align*}
which allow us to write:
\begin{equation*}
\frac{\e^{\Delta_w\tau}-1}{\Delta_w}-\frac{\e^{\Delta_a\tau}-1}{\Delta_a}\approx\frac{\left(\Delta_w-\Delta_a\right)\tau^2}{2}.
\end{equation*}
As a result, the integrand in \eqref{meantimet2} can be written as:
\begin{align*}
&\exp\left[-\frac{uvNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_w\tau}-1}{\Delta_w}-\frac{\e^{\Delta_a\tau}-1}{\Delta_a}\right)-vNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}\right]\approx\exp\left(-uvNp_m\tau^2-vNp_m\tau\right)\\
&=\exp\left(\frac{vNp_m}{2}\right)\exp\left[-\frac{uvNp_m}{2}\left(\tau+\frac{1}{u}\right)\right].
\end{align*}
Consequently, the mean time $\tau_2$ is obtained to be:
\begin{align}\label{limitapprox2}
\tau_2\approx\left[1+\exp\left(-Np_w\right)\right]\exp\left(\frac{vNp_m}{2u}\right)\frac{\erfc\left(\sqrt{\frac{vNp_m}{2u}}\right)}{\sqrt{\frac{2uvNp_m}{\pi}}},
\end{align}
where $\erfc$ is the complementary error function. We plot the expansion \eqref{limitapprox2} in \Cref{MeanTimeGrowthAneuploidyPlot} and observe that it is a very good fit for large values of the initial wildtype population size.

If we select only linear terms in the following expansions:
\begin{align*}
\frac{\e^{\Delta_w\tau}-1}{\Delta_w}&=\frac{1+\Delta_w\tau+O(\tau^2)-1}{\Delta_w}=\tau+O(\tau^2),\\
\frac{\e^{\Delta_a\tau}-1}{\Delta_a}&=\frac{1+\Delta_a\tau+O(\tau^2)-1}{\Delta_a}=\tau+O(\tau^2),
\end{align*}
we obtain the first order approximation for $\tau_2$:
\begin{align}\label{limitapprox3}
\tau_2\approx\left(1+\e^{-Np_w}\right)\int_0^\infty\e^{-uNp_m\tau}\,\d\tau=\frac{\left(1+\e^{-Np_w}\right)}{uNp_m} ,
\end{align}
which we plot in \Cref{MeanTimeGrowthAneuploidyPlot} and observe that it offer as a good a fit to \cref{meantimet2} as \cref{limitapprox2}. Additionally, we observe that for large initial wildtype populations sizes direct mutation drives evolutionary rescue while aneuploidy plays a role for intermediate sized tumors. This is consistent with the information obtained from \cref{rescue_ratio} where aneuploidy improves the probability of evolutionary rescue only  for small and intermediate values of $N$.

@@

\begin{figure}
% small population size
% switch from tolerance to non-growing
% TODO why no switch in panel B?
% TODO more simulations for 0 > Delta_a > Delta^*_a
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=1\textwidth]{Figures/P_est_divergence.pdf}
\end{subfigure}
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=1\textwidth]{Figures/P_est_divergenceLargePopulation.pdf}
\end{subfigure}
\caption{\textbf{Evolutionary rescue probability with tolerant or non-growing aneuploid cells.}
Rescue probability grows with the aneuploid growth rate $\Delta_a$ (x-axis), and is much higher in an initially large tumor than in a small one (\textbf(A) $N=10^4$; \textbf{(B)} $N=10^8$). 
Markers for simulation results with 95\% CI; 
solid lines for the exact formula (\cref{survival_prob} in \cref{rescue_prob});
dashed lines for the approximate formula (\cref{rescue_prob_approx}).
The approximation agrees with the simulation and exact solution when the initial tumor size is large (panel B).
When the tumor size is small (panel A), we switch between the approximation for tolerant and for non-growing aneuploid cells; the switch occurs at $\Delta_a^*=2vp_m+v+2\sqrt{vp_m\left(vp_m+\mu_a+v\right)}$.
Parameters: $\lambda_w=\lambda_a=\lambda_m=0.14$; $\mu_m=0.13$; $u=10^{-2}$; $v=10^{-7}$.
}
\label{rescue_prob_an_growth}
\end{figure}

>>>>

Even when evolutionary rescue occurs, it may take a long time.
We therefore wish to estimate the mean waiting time for rescue and the effect aneuploidy may have on it. 
We calculate the mean time for the appearance of the first mutant that rescues the cell population.
This can occur either through the evolutionary trajectory $wildtype \rightarrow mutant$ or through the trajectory $wildtype \rightarrow aneuploid \rightarrow mutant$.
We start with the former.

Assuming no aneuploidy ($u=0$), we define $T_1$ to be the time at which the first mutant cell appears that will avoid extinction and will therefore rescue the population.
Note that if extinction occurs, that is the frequency of mutants after a very long time is zero, $m_{\infty}=0$, then it is implied that $T_1=\infty$, and vice versa if $T_1<\infty$ then $m_{\infty}>0$.

The number of successful mutants generated until time $t$ can be approximated by an inhomogeneous Poisson process with rate $R\left(t\right) = u p_a w_t$,
where $w_t=N\e^{\Delta_w t}$ is the number of wildtype cells at time $t$.
Note that 
\begin{equation}\label{integralR}
\int_0^{t}{R(z)\d z} = 
u p_a N \frac{\exp[{\Delta_w t}]-1}{\Delta_w} \approx 
u p_a N t,
\end{equation}
by integrating the exponential and because $\frac{\exp[\Delta_w t]-1}{\Delta_w}=\frac{1+\Delta_w t+O(t^2)-1}{\Delta_w}=t+O(t^2)$.
The probability density function of $T_1$ is thus
$R\left(t\right)\exp\left(-\int_0^{t}{R(z)\d z}\right)$. % TODO ref
Therefore, the probability density function of the conditional random variable $(T_1 \mid T_1 < \infty)$ is
$f_1(t) = \frac{R\left(t\right)\exp\left(-\int_0^{t}{R(z)\d z}\right)}{p_{rescue}}$. 
\\

We are interested in the mean conditional time, $\tau_1=\mathbb{E}\left[T_1 \mid T_1<\infty\right]$, which is given by
\begin{equation}\label{meantime1}
\begin{aligned}
\tau_1 =
\int_{0}^{\infty}{t f_1(t) \d t} = 
\frac{\int_{0}^{\infty}{tR(t)\exp\left(-\int_0^{t}{R(z)\d z}\right) \d t}}{p_{rescue}} = 
\frac{\int_{0}^{\infty}{\exp\left(-\int_0^{t}{R(z)\d z}\right) \d t}}{p_{rescue}}
\end{aligned}
\end{equation}
after applying integration by parts.
Therefore, plugging \cref{integralR,rescue_prob2} in \cref{meantime1}, 
\begin{align}\label{limitapprox}
\tau_1 = 
%\frac{\int_0^\infty\exp\left(-\int_0^\tau R(z)\d z\right)\d t}{1-\left(1-p_w\right)^N} = 
\frac{\int_0^\infty\e^{-uNp_a\frac{\e^{\Delta_w\tau}-1}{\Delta_w}}\,\d\tau}{1-\left(1-p_w\right)^N} \approx
\frac{\int_{0}^{\infty}{\exp\left(-u p_a N t\right) \d t}}{1-\e^{-Np_w}} \approx \\
\left(1+\e^{-Np_w}\right)\int_0^\infty\e^{-u p_a N t} \d t =
\frac{1+\e^{-Np_w}}{u p_a N} ,
\end{align}
where we use the approximations 
$\frac{\e^{\Delta_w\tau}-1}{\Delta_w}=\frac{1+\Delta_w\tau+O(\tau^2)-1}{\Delta_w}=\tau+O(\tau^2)$ and $(1-\e^{-Np_w})^{-1} \approx 1+\e^{-Np_w}$ and integrate the exponent.
\Cref{MeanTimeGrowthAneuploidyPlot}B show the agreement between this approximating and simulation results for intermediate and large tumor sizes.
\\

When $Nu\gg1$ the aneuploid frequency dynamics is roughly deterministic and therefore can be approximated by 
\begin{equation}
a_t \approx \frac{Nu\e^{\Delta_wt}}{\Delta_w-\Delta_a}\left[1-\e^{\left(\Delta_w-\Delta_a\right)t}\right].
\end{equation}
As a result, when $N\gg1$ % TODO N>>1 ? or Nu >> 1?
the number of successful mutants created by direct mutation and via aneuploidy can be approximated by inhomogeneous Poisson processes with the rates
\begin{align}
r_1\left(t\right)&=vp_m\int_0^ta_{\tau}\,\d\tau=\frac{uvNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_wt}-1}{\Delta_w}-\frac{\e^{\Delta_at}-1}{\Delta_a}\right),\\
r_2\left(t\right)&=vp_m\int_0^tw_{\tau}\,\d\tau=vNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}.
\end{align}
For large initial population sizes we assume that the two processes are independent and as a result, they can be merged into a single Poisson process with rate $\left(r_1+r_2\right)\left(t\right)$.
Consequently, the mean time to the appearance of the first rescue mutant is
\begin{align}\label{meantimet2}
\tau_2 = 
\frac{\int_0^\infty\e^{-\left(r_1+r_2\right)} \d t}{1-\left(1-p_w\right)^N} = 
\frac{\int_0^\infty\exp\left[-\frac{uvNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_w\tau}-1}{\Delta_w}-\frac{\e^{\Delta_a\tau}-1}{\Delta_a}\right)-vNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}\right] \d t}{1-\left(1-p_w\right)^N},
\end{align}
which we plot in \Cref{MeanTimeGrowthAneuploidyPlot}A as a function of the initial population size, $N$.

We wish to obtain a simpler formula for $\tau_2$, similar to \cref{limitapprox}.
We thus have the following expansions,
%\begin{align*}
%\frac{\e^{\Delta_w\tau}-1}{\Delta_w}&=\frac{1+\Delta_w\tau+\frac{\Delta_w^2\tau^2}{2}+O(\tau^3)-1}{\Delta_w}=\tau+\frac{\Delta_w}{2}\tau^2+O(\tau^3) ,\\
%\frac{\e^{\Delta_a\tau}-1}{\Delta_a}&=\frac{1+\Delta_a\tau+\frac{\Delta_a^2\tau^2}{2}+O(\tau^3)-1}{\Delta_a}=\tau+\frac{\Delta_w}{2}\tau^2+O(\tau^3) ,
%\end{align*}
%which allow us to write
%\begin{equation*}
%\frac{\e^{\Delta_w\tau}-1}{\Delta_w}-\frac{\e^{\Delta_a\tau}-1}{\Delta_a} \approx \frac{\left(\Delta_w-\Delta_a\right)\tau^2}{2}.
%\end{equation*}
%As a result, the integrand in \cref{meantimet2} can be written as:
%\begin{align*}
%&\exp\left[-\frac{uvNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_w\tau}-1}{\Delta_w}-\frac{\e^{\Delta_a\tau}-1}{\Delta_a}\right)-vNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}\right]\approx \\
%& \exp\left(-uvNp_m\tau^2-vNp_m\tau\right) =
%\exp\left(\frac{vNp_m}{2}\right)\exp\left[-\frac{uvNp_m}{2}\left(\tau+\frac{1}{u}\right)\right].
%\end{align*}
%Consequently, the mean conditional time $\tau_2$ is approximated by
%\begin{align}\label{limitapprox2}
%\tau_2 \approx
%\frac{\left(1+\exp\left(-Np_w\right)\right) \exp\left(\frac{vNp_m}{2u}\right) \erfc\left(\sqrt{\frac{vNp_m}{2u}}\right)}{\sqrt{\frac{2uvNp_m}{\pi}}},
%\end{align}
%where $\erfc(z)$ is the complementary error function. 
%
%If we apply linear approximations,
\begin{align*}
\frac{\e^{\Delta_w\tau}-1}{\Delta_w}&=\frac{1+\Delta_w\tau+O(\tau^2)-1}{\Delta_w}=\tau+O(\tau^2),\\
\frac{\e^{\Delta_a\tau}-1}{\Delta_a}&=\frac{1+\Delta_a\tau+O(\tau^2)-1}{\Delta_a}=\tau+O(\tau^2),
\end{align*}
which we use to derive a first-order approximation for $\tau_2$,
\begin{align}\label{limitapprox3}
\tau_2 \approx
\left(1+\e^{-Np_w}\right)\int_0^\infty\e^{-uNp_m\tau}\,\d\tau=\frac{\left(1+\e^{-Np_w}\right)}{uNp_m} ,
\end{align}

\Cref{MeanTimeGrowthAneuploidyPlot}A shows that the approximation \cref{limitapprox3} has a good fit with simulation results for large initial wildtype population size ($N > 2\cdot10^7$). An approximation that uses second-order terms $\frac{\Delta_w^2\tau^2}{2}$ and $\frac{\Delta_a^2\tau^2}{2}$ does not perform better (\Cref{MeanTimeGrowthAneuploidyPlot}A).

% TODO YOAV HERE

Additionally, we observe that for large initial wildtype populations sizes direct mutation drives evolutionary rescue while aneuploidy plays a role for intermediate sized tumors. This is consistent with the information obtained from \cref{rescue_ratio} where aneuploidy improves the probability of evolutionary rescue only  for small and intermediate values of $N$.

@@

\begin{figure}
% small population size
% switch from tolerance to non-growing
% TODO why no switch in panel B?
% TODO more simulations for 0 > Delta_a > Delta^*_a
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=1\textwidth]{Figures/P_est_divergence.pdf}
\end{subfigure}
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=1\textwidth]{Figures/P_est_divergenceLargePopulation.pdf}
\end{subfigure}
\caption{\textbf{Evolutionary rescue probability with tolerant or non-growing aneuploid cells.}
Rescue probability grows with the aneuploid growth rate $\Delta_a$ (x-axis), and is much higher in an initially large tumor than in a small one (\textbf{A}: $N=10^4$; \textbf{B}: $N=10^8$). 
Markers for simulation results with 95\% CI; 
solid lines for the exact formula (\cref{survival_prob} in \cref{rescue_prob});
dashed lines for the approximate formula (\cref{rescue_prob_approx}).
The approximation agrees with the simulation and exact solution when the initial tumor size is large (panel B).
When the tumor size is small (panel A), we switch between the approximation for tolerant and for non-growing aneuploid cells; the switch occurs at $\Delta_a^*=2vp_m+v+2\sqrt{vp_m\left(vp_m+\mu_a+v\right)}$.
Parameters: $\lambda_w=\lambda_a=\lambda_m=0.14$; $\mu_m=0.13$; $u=10^{-2}$; $v=10^{-7}$.
}
\label{rescue_prob_an_growth}
\end{figure}
